---
title: "individual1"
author: "Maddie Mueller"
date: "4/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab 5
# Individual 1 - Madalyn Mueller

```{r}
#All defaults
library(tidyverse)
```

```{r}
#All defaults
acc <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)

per <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)
```

## 1) Are there some days of the week where more accidents happen than the others (see FARS manual, use variable DAY_WEEK)?

```{r}
#All defaults
ggplot(acc, aes(x=DAY_WEEK)) + geom_bar()
```

Yes, according to this chart the most accidents happen on Saturdays, followed by Fridays and then Sundays. This makes sense, because the weekend is when people typically drink, act reckless, and travel.

## 2) Create a data frame containing the persons who are fatally hurt in the accidents (look up variable INJ_SEV)

```{r}
#All defaults
fatal_inj <- per %>%
  filter(INJ_SEV == 4)
print(fatal_inj)
```

## 3) Create a data frame containing the most dangerous vehicle make in each state. The number of persons fatally hit in the vehicle make is used to assess the (non-)safety of a make. Make sure to handle the missing values appropriately. (look up variable MAKE)


```{r}
#All defaults
danger_car <- fatal_inj %>%
  drop_na(MAKE) %>%
  group_by(STATE, MAKE) %>%
  summarize(count = n()) %>%
  group_by(STATE) %>%
  filter(count==max(count))

danger_car
```

## 4) Create a map, and label each state with the most dangerous vehicle. Discuss the definition of the most dangerous vehicle, and what you find from the map. (Hint: Read the description for the STATE and COUNTY columns in the FARS manual. The state & county codes are Geographic Locator Codes (GLCs) from the General Services Administration’s (GSA) publication. Use readxl::read_xlsx to read in the GLCs.)

```{r}
#All defaults
accper <- inner_join(acc, per, by ="ST_CASE")
fatal_accper <- accper %>%
  filter(INJ_SEV == 4)
ggplot(fatal_accper, aes(x=LONGITUD, y=LATITUDE, fill=MAKE)) + geom_polygon(aes(group=STATE.x))
```

## 5) Join the accident and person table (work out which variable(s) to use)

```{r}
#All defaults
accper <- inner_join(acc, per, by ="ST_CASE")
```

## 6) Tally the number of accidents by day of the week (DAY_WEEK), hour of the day (HOUR) and gender (SEX). Visualize the results and explain what you find.

```{r}
#All defaults
acc %>%
  group_by(DAY_WEEK) %>%
  count(DAY_WEEK)
```
It looks like the most accidents happen on Saturdays, followed by Sundays and Fridays. The least amount of accidents happen on Wednesdays, which makes sense because it is in the middle of the week when there usually isn't a lot going on.

```{r}
#All defaults
acc %>%
  group_by(HOUR) %>%
  count(HOUR)
```
It looks like the most accidents happened around 6:00 p.m. This makes sense because that is when most people are coming home from work on the weekdays or going out to dinner/drinks/events on the weekends.

```{r}
#All defaults
per %>%
  group_by(SEX) %>%
  count(SEX)
```
Males have twice the amount of accidents that females have. This is surprising to me, I figured the numbers would be more even. Males must be more of reckless drivers.

## 7) Now plot a choropleth map of the number of deaths on a county level. Also explain what you find.

```{r}
#All defaults
fatal_accper <- accper %>%
  filter(INJ_SEV == 4)
ggplot(fatal_accper, aes(x=LONGITUD, y=LATITUDE, fill=INJ_SEV)) + geom_polygon(aes(group=COUNTY.x))
```

## 8) Is summer or winter more dangerous? Does this depend on states? Explore and explain.

```{r}
#All defaults
summerdeath <- acc %>%
  filter(MONTH >= 6 | MONTH <= 8) %>%
  select(STATE, FATALS) %>%
  group_by(STATE) %>%
  summarize(sumCount = sum(FATALS))

winterdeath <- acc %>%
  filter(MONTH >= 11 | MONTH <= 2) %>%
  select(STATE, FATALS) %>%
  group_by(STATE) %>%
  summarize(winCount = sum(FATALS))

summerdeath
winterdeath

totalsummer <- sum(summerdeath$sumCount)
totalsummer

totalwinter <- sum(winterdeath$winCount)
totalwinter
```
There are more car accidents in the summer than there are in the winter. This is pretty consistent throughout all the states. This makes sense because there is no school during the summer, so there are more people going out and doing fun things. But it still surprised me, because I thought winter would have more deaths because of the weather conditions of states with changing seasons.



