---
title: 'Lab #5: Drunk Driving & Fatal Accidents'
author: "Sydney Thompson"
date: "4/13/2021"
output: html_document
---
# FARS Data

```{r}
acc <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/accident.csv", stringsAsFactors = FALSE)
```

```{r}
pers <- read.csv("https://raw.githubusercontent.com/xdaiISU/ds202materials/master/hwlabs/fars2017/person.csv", stringsAsFactors = FALSE)
```

```{r}
library(lubridate)
library(tidyverse)
library(ggplot2)
```

# Questions

# 1) Are there some days of the week where more accidents happen than the others (see FARS manual, use variable DAY_WEEK)?

```{r}
ggplot(acc, aes(x=DAY_WEEK))+geom_bar()+xlab('Days of the Week')+ylab('Number of Accidents')+ggtitle('Number of Accidents for Each Day of the Week')
```

Yes. More accidents occur on the weekends, Friday, Saturday and Sunday. This is usually due to people not having work on the weekends which allows for people to drink and/or travel more on these days making accidents more prone. 

# 2) Create a data frame containing the persons who are fatally hurt in the accidents (look up variable INJ_SEV)

```{r}
fatal_inj <- pers %>% group_by(INJ_SEV) %>% filter(INJ_SEV == 4)
fatal_inj
```

# 3) Create a data frame containing the most dangerous vehicle make in each state. The number of persons fatally hit in the vehicle make is used to assess the (non-)safety of a make. Make sure to handle the missing values appropriately. (look up variable MAKE)

```{r}
dang_vehicle <- fatal_inj %>% drop_na(MAKE)%>% group_by(STATE, MAKE) %>% summarize(count=n()) %>% group_by(STATE) %>% filter(count == max(count))
dang_vehicle
```

# 4) Create a map, and label each state with the most dangerous vehicle. Discuss the definition of the most dangerous vehicle, and what you find from the map. (Hint: Read the description for the STATE and COUNTY columns in the FARS manual. The state & county codes are Geographic Locator Codes (GLCs) from the General Services Administration’s (GSA) publication. Use readxl::read_xlsx to read in the GLCs.)

```{r}
states <- map_data('state')
str(states)
```

```{r}
acc_pers <- inner_join(acc, pers, by ="ST_CASE")
fatal_acc_pers <- acc_pers %>% filter(INJ_SEV == 4)
ggplot(fatal_acc_pers, aes(x=LONGITUD, y=LATITUDE, fill=MAKE)) + geom_polygon(aes(group=STATE.x))
```

# 5) Join the accident and person table (work out which variable(s) to use)

```{r}
acc_pers <- inner_join(acc, pers, by ="ST_CASE")
acc_pers
```

# 6) Tally the number of accidents by day of the week (DAY_WEEK), hour of the day (HOUR) and gender (SEX). Visualize the results and explain what you find.

```{r}
acc_day <- acc %>% group_by(DAY_WEEK) %>% count(DAY_WEEK)
acc_day
```

For accidents by day of the week, one can see that days Friday(6), Saturday(7) and Sunday(1) have more accidents occur due to more travel on the weekends as people have free time off of work/school. 

```{r}
acc_hour <- acc %>% group_by(HOUR) %>% count(HOUR)
acc_hour
```

For accidents per hour of the day, more accidents occur during 5-6am for morning rush hour to work and around 6pm when individuals are traveling home from work, hauling children to sports, going out to eat, etc. 

```{r}
acc_gender <- pers %>% group_by(SEX) %>% count(SEX)
acc_gender
```

For accidents based on gender, it appears that men(1) are in double the amount of accidents than women(2). This is surprising to me, but men seem to be more reckless at driving than women. 

# 7) Now plot a choropleth map of the number of deaths on a county level. Also explain what you find.

```{r}
acc_pers_deaths <- acc_pers %>% filter(INJ_SEV == 4)
ggplot(acc_pers_deaths, aes(x = LONGITUD, y = LATITUDE, fill = INJ_SEV))+geom_polygon(aes(group=COUNTY.x))
```

# 8) Is summer or winter more dangerous? Does this depend on states? Explore and explain.

```{r}
summer <- acc %>% filter(MONTH >= 6 | MONTH <= 8) %>% select(STATE, FATALS) %>% group_by(STATE) %>% summarize(sumCount = sum(FATALS))
summer
winter <- acc %>% filter(MONTH >= 11 | MONTH <= 2) %>% select(STATE, FATALS) %>% group_by(STATE) %>% summarize(winCount = sum(FATALS))
winter
total_summer_fatal <- sum(summer$sumCount)
total_summer_fatal
total_winter_fatal <- sum(winter$winCount)
total_winter_fatal
```
From the data shown above, the total summer fatalities is 37,133 and the total winter fatalities is 11,496. With that being said, summer is more dangerous for driving a vehicle around rather than in the winter. I believe this is due to kids being out of school during the summer and driving more places to keep busy. Although, this did surprise me as I had assumed winter would have more fatalities due to harsh weather conditions. I also noticed that the fatalities did not play a huge role depending on the state. 